|

### 2021-04-13 02-Chp-Background.Rmd
End of day-2592
End of day: -2405

---
output:
  bookdown::word_document2: default
  bookdown::pdf_document2:
    template: templates/brief_template.tex
  bookdown::html_document2: default
documentclass: book
---

@@

```{r, echo = FALSE, warning=FALSE, message=FALSE}
source("R/doc_options.R")
source("R/helper.R")
knitr::read_chunk("R/background.R")

knitr::opts_chunk$set(
  include = FALSE
)
```

@@

<!-- ## Lay summary -->

<!-- Around 850,000 people in the UK live with dementia, and by 2040, nearly twice as many will have the condition. Despite many promising candidates, no cure for dementia currently exists, so the focus is now on figuring out ways to prevent the condition by finding risk factors (characteristics that influence a person's chance of developing a disease) that we can easily change. Avoiding a risk factor does not guarantee that a person will not develop dementia, but makes it less likely. A key risk factor for dementia may be the levels of lipids (fatty substances such as cholesterol) in a person's blood.  -->

<!-- This Chapter provides background information on dementia and lipids. It also provides a description. -->

<!-- Evidence synthesis involves drawing on multiple sources of information to obtain the best possible to a research question, in this case, do lipids influence dementia risk.  -->

@@

This Chapter provides an overview of the broad context of this thesis, introducing the core concepts used throughout and providing some background context on each. It will briefly discuss the underlying pathologies and diagnosis of dementia, its public health importance, and the current state of treatment and prevention therapeutics.

In this context, it will highlight the importance of identifying easily modified risk factors and introduce blood lipids, and blood lipid-modifying treatments such as statins, as the primary exposures considered by this research.

The central theoretical framework used - the synthesis of diverse sources of evidence - is introduced, and the research presented in this thesis is then framed in terms of three types of diverse evidence, with opposing viewpoints for each being discussed.

@@

Dementia is a collective umbrella term for a range of conditions each with different clinical presentations and disease courses, and is generally defined as "". <!--- Cite --->

Alzheimer's disease, named after Dr. Alois Alzheimer who presented the first recorded case in 1907, is the most common cause of dementia, accounting for approximately 60-80% of cases. Characterised by substantial cognitive impairment and diffculty with high level executive function to the extent that it interferes with, it is an insidious disease, within initial onset thought to occur up to 15 years prior to symptomatic presentation. Much remains unknown unknown about Alzheimer's pathogenesis, despite research implicating the accumluation of amyloid plaques and neurofibrillary tangles, or alternatively the cholingeric nervous system, <!--- check ---> as potential mechanisms of disease.

@@

Vascular dementia (VaD) is the second largest underlying pathology of dementia, accounting for ~10% of cases and is caused by a range of cerebrovascular disorders. Presentation can vary widely based on the underlying cause, and due to the varied underlying  pathopysiology, VaD can onset quite rapidly following a cerbrovascular event such as a stroke or over a long period due to series of small infarcts. VaD is diagnosed using the NINCDS-AIREN criteria.[@roman1993vascular]

@@

Where more than one underlying dementia pathologies co-occur, this is termed mixed dementia. It is common for patients to present with mixed dementia, as the occurrence of one underlying cause weakens overall brain health and increases the likeliness of occurrence of the other. <!--- Double check this --->

@@

Dementia is difficult to diagnose, primarily due to the absence of a gold standard test for the condition. <!-- is this true, or would it be better to say that there is no gold standard test without an autopsy?--> There exist several diagnostic criteria for dementia and related disease, many of which have existed for quite some time and have more recently been updated to reflect current understanding in dementia pathology. In all cases, dementia is diagnosed on the basis of behavioural and congitive changes as assessed by an experience clinician. 
<!-- This means the potential for misclassification both of dementia versus no-dementia and between the different underlying conditions is extremely likely. -->

@@

Two of the most commonly used criteria include the the Diagnostic and Statistical Manual of Mental Disorders, fourth edition (DSM-IV-TR) <!-- __[CITATION NEEDED]__ --> criteria and the National Institute of Neurological Disorders and Stroke--Alzheimer Disease and Related Disorders (NINCDS-ADRDA) criteria.[@dubois2007]

@@

### Public Health Importance

@@

Total serum cholesterol is a commonly used summary measure to estimate the total lipid burden in a patients blood, and  derived from the HDL-c, LDL-c and TG fractions using the Friedewald formula:[@friedewald1972]

$$TC \approx LDLc - HDLc - kTG$$ where $k$ is 0.20 if measurements are in mg/dl and 0.45 if in mmol/l.

<!--- Include formula for working out total cholesterol and a table detailing the acceptable ranges. --->

Elevated cholesterol in the bloodstream via high LDL-c levels leads to atherosclerosis, the build-up of fatty deposits in the blood vessels, which constrict blood flow and can lead to vascular complications. Alternatively, part of the deposit can detach from the artery walls, forming a clot that can lead to a heart attach or stroke. Globally, the prevalence of elevated total cholesterol, or hypercholestoreolemia, was estimated by the World Health Organisation to be approximately 40%.

@@

Statins are by far the most common method of lipid regulation currently prescribed. Statins inhibit the conversion of 3-Hydroxy-3-Methylglutaryl-Coenzyme-A (HMG-CoA) into mevalonic acid,  by compeitivlely binding with HMG-CoA Reductase (HMG-CoA-R). This conversion is the key rate limiting step in the biochemical pathway that produces cholesterol (see Figure \ref(), allowing statins to reduce the production of LDL cholesterol.

@@

There are a range of other interventions that can be used to modify a persons lipid profile, though each act is slightly different ways. Most commonly though, these treatments are either used as adjunct (additional) treatments with statins therapy, or are used in situations where statins are contra-indicated or not tolerated.

@@

The most commonly used non-statin therapeutic is ezetimbibe, which prevents intestinal absorption of cholesterol. However, when used alone, it has a limited LDL-c lowering effect, leading to the creation of combined statin/ezetimbe therapies (both compounds contained in a single pill, as opposed to complimentary treatemnts).

@@

@@

In this section I provide an overview of the varying sources of evidence on the relationship between blood lipid levels and dementia risk.

@@

A role for lipids in the aetiology of dementia is supported by both genetic linkage studies and functional cell biology studies. The generation of the amyloid plaques found in the brains of Alzheimer's patients is cholesterol dependent [CITE], while the most established genetic risk factor for late-onset dementia, apolipoprotein E (ApoE), is involved in cerebral cholesterol transport. Several other genes involved in cholesterol transport have also been found to be robustly associated with increased AD susceptibility [CITE].

@@

Despite these encouraging results, evidence from the diverse range of epidemiological studies on this topic has to date been been contradictory.

@@

Several observational studies have examined the relationships between concentrations of serum lipids (total cholesterol (TC), low density lipoprotein cholesterol (LDL-c), high density lipoprotein cholesterol (HDL-c) and triglycerides) and both Alzheimer's disease and vascular dementia and reported widely different results. A high serum cholesterol concentration has been found to be associated with an increase in susceptibility to AD [8-13], though other studies have shown no association [14-17], or a reduced susceptibility [18,19]. With regards VaD, decreased levels of HDL-c appear to be associated with increased risk [18,20,21], while for LDL-c, studies have reported both positive and negative association [18,22].

@@

### Summary level results vs individual patient data

@@

## Aims and Objectives of the thesis

@@

### Aims and Objectives

@@

### Thesis Structure

>>>>

---
bibliography: bibliography/references.bib
csl: bibliography/nature.csl
output:
  bookdown::pdf_document2:
    template: templates/brief_template.tex
  bookdown::word_document2: default
  bookdown::html_document2: default
documentclass: book
---

@@

```{r, echo = FALSE, warning=FALSE, message=FALSE}
source("R/doc_options.R")
source("R/helper.R")
knitr::read_chunk("R/background.R")
doc_type <- knitr::opts_knit$get('rmarkdown.pandoc.to') # Info on knitting format
```

@@

## Figures/Tables To

* Diagnostic criteria
* Statin mechanism of action 
* Evidence pyramid
* Overview of MR studies

## Lay summary

Around 850,000 people in the UK live with dementia, and by 2040, nearly twice as many will have the condition. Despite many promising candidates, no cure for dementia currently exists, meaning the focus is on finding ways to prevent the condition. The best way to do this is to find risk factors (characteristics that influence a person's chance of developing a disease) for dementia that we can easily change. Avoiding a risk factor does not guarantee that a person will not develop dementia, but makes it less likely. A key risk factor for dementia may be the levels of lipids (fatty substances such as cholesterol) in a person's blood, though not all available research agrees. The aim of this thesis is to use all available evidence to assess whether blood lipids levels are in fact a risk factor for dementia.

This introductory chapter provides background information on both dementia and blood lipids, and on the potential link between them. It introduces the combination of evidence from different sources as the key framework used to guide the research presented in the remaining chapters. It then maps the formal aims and objectives of the research project to the relevant chapters of this thesis. Finally, it summarises the outputs (journal articles, presentations and software) that were created as part of this thesis.

@@

This Chapter provides an overview of the broad context of this thesis, introducing the core concepts used throughout and providing some background on each. It will briefly discuss the underlying pathologies and diagnosis of dementia, its public health importance, and the current state of treatment and prevention therapeutics.

This chapter will highlight the importance of identifying easily modified risk factors and introduce blood lipids, and blood lipid-modifying treatments such as statins, as the primary exposures considered by this research.

The central theoretical framework used throughout this thesis will be the synthesis of diverse sources of evidence. This framework is introduced, and the research presented in this thesis is then framed in terms of three types of diverse evidence, with opposing viewpoints for each being discussed.

@@

Dementia is a collective umbrella term for a range of conditions each with different clinical presentations and disease courses, and is generally __defined as ""__. <!--- Cite --->

Alzheimer's disease, named after Dr. Alois Alzheimer who presented the first recorded case in 1907, is the most common cause of dementia, accounting for approximately 60-80% of cases. Characterised by substantial cognitive impairment and diffculty with high level executive function to the extent that it interferes with, it is an insidious disease, within initial onset thought to occur up to 15 years prior to symptomatic presentation. Despite research implicating the accumulation of amyloid plaques and neurofibrillary tangles, or alternatively the cholingeric nervous system, <!--- check ---> as potential mechanisms of disease, much remains unknown unknown about Alzheimer's pathogenesi

@@

Vascular dementia (VaD) is the second largest underlying pathology of dementia, accounting for ~10% of cases and is caused by a range of cerebrovascular disorders. Presentation can vary widely based on the underlying cause. Due to the varied underlying  pathophysiology, VaD can either onset quite rapidly following a cerebrovascular event such as a stroke or over a long time-frame due to a series of small infarcts. VaD is diagnosed using the __NINCDS-AIREN__ criteria.[@roman1993vascular]

@@

The term mixed dementia is used when more than one underlying dementia pathologies co-occur. This is a  common presentation of dementia because the presence  of one underlying condition  weakens overall brain health and increases the likelihood  of another condition developing. <!--- Double check this --->

@@

Dementia is difficult to diagnose, primarily due to the absence of a gold standard test for the condition. <!-- is this true, or would it be better to say that there is no gold standard test without an autopsy?--> There exist several diagnostic criteria for dementia and related disease. In all cases, dementia is diagnosed on the basis of behavioural and cognitive changes as assessed by an experience clinician. 
<!-- This means the potential for misclassification both of dementia versus no-dementia and between the different underlying conditions is extremely likely. -->

@@

Two of the most commonly used criteria include those found in the Diagnostic and Statistical Manual of Mental Disorders, fourth edition (DSM-IV-TR) <!-- __[CITATION NEEDED]__ --> criteria and the National Institute of Neurological Disorders and those developed by the Stroke-Alzheimer's Disease and Related Disorders Association (NINCDS-ADRDA).[@dubois2007]

@@

### Public health importance

@@

Total serum cholesterol (TC) is a commonly-used summary measure to estimate the total amount of lipid present in the blood. The measure is derived from measurements of the individual HDL-c, LDL-c and TG levels using the Friedwald formula:[@friedewald1972]

$$TC \approx LDLc - HDLc - kTG$$
where $k$ is 0.20 if measurements are in milligrams per decilitre (_mg/dl_) and 0.45 if measured in millimole per litre (_mmol/l_).

Widely used criteria for assessing the acceptable levels of different types of lipids are based on the National Cholesterol Education Program (NCEP)[@national2002third], and are outlined in Table \@ref(tab:lipidLevels-table).

(ref:lipidLevels-caption) Classificaion of lipid levels according to the National Cholesterol Education Program guidelines.[@national2002third]

```{r lipidLevels-table, message=FALSE, results="asis", echo = FALSE}
```
<!----------------------------------------------------------------------------->

Elevated cholesterol in the bloodstream via high LDL-c levels leads to atherosclerosis, the build-up of fatty deposits in the blood vessels. These deposits constrict blood flow and can lead to vascular complications. Alternatively, part of the deposit can detach from the artery walls, forming a clot that can lead to a heart attach or stroke. Globally, the prevalence of elevated total cholesterol, or hypercholesterolemia, was estimated by the World Health Organisation to be approximately 40%.

@@

Statins are by far the most common method of lipid regulation currently prescribed. Statins inhibit the conversion of 3-Hydroxy-3-Methylglutaryl-Coenzyme-A (HMG-CoA) into mevalonic acid,  by competitively binding with HMG-CoA Reductase (HMG-CoA-R). This conversion limits the rate at which cholesterol (see Figure \ref(), can be produced, enabling statins to effectively reduce the production of LDL cholesterol.

@@

There are a range of other interventions that can be used to modify a persons lipid profile, which each acting in slightly different ways. However, in general, these treatments are either used as adjunct (additional) treatments with statins therapy or are used in situations where statins are contra-indicated or not tolerated.

@@

The most commonly used non-statin therapeutic is ezetimibe, which prevents intestinal absorption of cholesterol. However, when used alone, it has a limited LDL-c lowering effect, leading to the creation of combined statin/ezetimibe therapies (both compounds contained in a single pill, as opposed to complimentary treatments).

@@

<!----------------------------------------------------------------------------->
(ref:lipidTreatments-caption) Summary of available treatments for hyperlipidaemia.

```{r lipidTreatments-table, message=FALSE, results="asis", echo = FALSE}
```
<!----------------------------------------------------------------------------->

@@

This section provides an overview of the varying sources of evidence on the relationship between blood lipid levels and dementia risk.

@@

A role for lipids in the aetiology of dementia is supported by both genetic linkage studies and functional cell biology studies. The generation of the amyloid plaques found in the brains of Alzheimer's patients is cholesterol dependent [CITE], while the most established genetic risk factor for late-onset dementia, apolipoprotein E (ApoE), is involved in cerebral cholesterol transport. Several other genes involved in cholesterol transport have also been found to be associated with increased AD susceptibility [CITE].

@@

Despite these results, evidence from the diverse range of epidemiological studies on this topic has been inconclusive.

@@

Several observational studies have examined the relationships between concentrations of serum lipids (total cholesterol (TC), low density lipoprotein cholesterol (LDL-c), high density lipoprotein cholesterol (HDL-c) and triglycerides) and both Alzheimer's disease and vascular dementia and reported extremely varied results. In some studies, a high serum cholesterol concentration has been found to be associated with an increase in susceptibility to AD [8-13], however others have shown no association [14-17], or a reduced susceptibility [18,19]. With regards VaD, decreased levels of HDL-c appear to be associated with increased risk [18,20,21], while for LDL-c, studies have reported both positive and negative association [18,22].

@@

### Summary level results versus individual patient data

@@

## Thesis overview

@@

### Aims and objectives

@@

### Structure

### 2021-04-13 03-Chp-Systematic-Review-Tools.Rmd
End of day: -2405

---
bibliography: bibliography/references.bib
csl: bibliography/bmj.csl
output:
  bookdown::pdf_document2:
    template: templates/brief_template.tex
  bookdown::word_document2: 
      toc: false
      toc_depth: 3
      reference_docx: templates/word-styles-reference-01.docx
  bookdown::html_document2: default
documentclass: book
---

>>>>

---
bibliography: bibliography/references.bib
csl: bibliography/nature.csl
output:
  bookdown::pdf_document2:
    template: templates/brief_template.tex
  bookdown::word_document2: 
      toc: false
      toc_depth: 3
      reference_docx: templates/word-styles-reference-01.docx
  bookdown::html_document2: default
documentclass: book
---

### 2021-04-13 index.Rmd
End of day: 0
End of day: -2405

#####################
## PDF formatting  ##
#####################
abstractseparate: true  # include front page w/ abstract for examination schools?
bib-humanities: false   #set to true if you want in-text references formatted as author-year
doi-in-bibliography: true #set to true if you want DOI's to be shown in the bibliography
draft: false # add as DRAFT mark in the footer?
page-layout: twoside #'nobind' for PDF output (equal margins), 'twoside' for two-sided binding (mirror margins and blank pages), leave blank for one-sided binding (left margin > right margin)
hidelinks: true #if false, the PDF output highlights clickable links with a colored border - you will probably want to set this to true for PDF version you wish to physically print
toc-depth: 2 # depth of heading to include in table of contents
lof: true # list of figures in front matter?
lot: true # list of tables in front matter?
mini-toc: true  # mini-table of contents at start of each chapter? (this just prepares it; you must also add \minitoc after the chapter titles)
mini-lot: false  # mini-list of tables by start of each chapter?
mini-lof: false  # mini-list of figures by start of each chapter?

@@

#####################
## output options  ##
#####################
output:
  bookdown::pdf_book:
    template: templates/template.tex
    keep_tex: true
    citation_package: default 
    pandoc_args: ["--lua-filter=scripts_and_filters/correction_filter.lua", "--csl","bibliography/nature.csl"] #remove filter to stop applying blue background to inline corrections
  bookdown::gitbook:
    css: templates/style.css
    download: ["pdf"]
    output_dir: "docs"
    split_bib: false
    config:
      toc:
        before: |
          <li><a href="./">Thesis - Draft</a></li>
        after: |
          <li><a href="https://github.com/mcguinlu/thesis" target="blank">Associated repository</a></li>
      sharing:
        facebook: false
        twitter: yes
        all: false
  bookdown::word_document2:
    output_dir: "docs"
    toc: true
    toc_depth: 2
    reference_docx: templates/word-styles-reference-01.docx   
link-citations: true
documentclass: book

>>>>

#####################
## PDF formatting  ##
#####################
abstractseparate: true  # include front page w/ abstract for examination schools?
bib-humanities: false   #set to true if you want in-text references formatted as author-year
doi-in-bibliography: true #set to true if you want DOI's to be shown in the bibliography
draft: false # add as DRAFT mark in the footer?
page-layout: twoside #'nobind' for PDF output (equal margins), 'twoside' for two-sided binding (mirror margins and blank pages), leave blank for one-sided binding (left margin > right margin)
hidelinks: true #if false, the PDF output highlights clickable links with a colored border - you will probably want to set this to true for PDF version you wish to physically print
toc-depth: 2 # depth of heading to include in table of contents
lof: true # list of figures in front matter?
lot: true # list of tables in front matter?
mini-toc: true  # mini-table of contents at start of each chapter? (this just prepares it; you must also add \minitoc after the chapter titles)
mini-toc-depth: 1
mini-lot: false  # mini-list of tables by start of each chapter?
mini-lof: false  # mini-list of figures by start of each chapter?

@@

#####################
## output options  ##
#####################
output:
  bookdown::pdf_book:
    template: templates/template.tex
    keep_tex: true
    citation_package: default 
    pandoc_args: ["--lua-filter=scripts_and_filters/correction_filter.lua", "--csl","bibliography/nature.csl"] #remove filter to stop applying blue background to inline corrections
  bookdown::gitbook:
    css: templates/style.css
    download: ["pdf"]
    output_dir: "docs"
    split_bib: false
    config:
      toc:
        before: |
          <li><a href="./">Thesis - Draft</a></li>
        after: |
          <li><a href="https://github.com/mcguinlu/thesis" target="blank">Associated repository</a></li>
      sharing:
        facebook: false
        twitter: yes
        all: false
  bookdown::word_document2:
    output_dir: "docs"
    toc: true
    toc_depth: 2
    reference_docx: templates/word-styles-reference-01.docx
    number_sections: false   
link-citations: true
documentclass: book
